Formalizing Stack Safety as a Security Property

### 2022-12-29 submission5/core.tex
Comments on the current draft, and some polishing here and there

``Stack safety'' is a term associated with a
variety of compiler, run-time, and hardware mechanisms for protecting stack
memory. But unlike heap memory, the stack does not correspond to any single
high-level language concept. The result is that
these mechanisms typically lack precise specifications,
relying instead on informal descriptions and examples of the bad
behaviors that they prevent. Such a specification is especially desirable
for an unusual enforcement mechanism, like the ``lazy'' stack safety
micro-policies proposed by Roessler and DeHon~\cite{DBLP:conf/sp/RoesslerD18}.

We propose a formal characterization of ``observable'' stack safety based on
concepts from language-based security. Stack safety is a
combination of an integrity property (``the private
state in each caller's stack frame is held invariant by the callee'')
and confidentiality properties (``the caller's and callee's behavior are
insensitive to one another's private state,''), both compatible with lazy tagging, as well as
control-flow property.

@@

The call stack is a perennial target for low-level attacks, leading to a
range of dire consequences, from leakage or corruption of private stack data
to control-flow hijacking. To foil such attacks, a profusion of
software and hardware protections have been proposed,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges+21},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18}.
  \ifaftersubmission\apt{Mostly from
  nick; there could be more}\bcp{Yes, going back to MIT days---we should
  include several more of these, if only to give readers the impression that
this is a well-studied mechanism (so formalizing its protections is
useful).}
\fi

The protections offered by such mechanisms are commonly described in terms
of concrete examples of attacks that they can prevent---corruption of return
addresses, buffer overflows, use of uninitialized variables, etc.---leaving
a more abstract characterization to the reader's intuition.  But these
mechanisms can be quite intricate, and the behaviors they aim to prevent are
subtle and varied. It can be hard to guess, based only on informal
intuitions, whether a given mechanism actually blocks all potential
attacks---or, conversely, whether it is overly conservative and disallows
system behaviors that are actually safe.
To settle such questions with confidence,
we need a precise, generic, and formal specification for stack
safety, as a basis both for comparing the security claims of different
enforcement techniques and for validating that these claims
are met by particular implementations. With the notable exception of the model used
in Skorstengaard et al.'s StkTokens work~\cite{SkorstengaardSTKJFP}, which we
discuss in Section \ref{sec:relwork}, there is a dearth of such properties in
the literature.

We propose a novel characterization of stack safety using the tools of language-based
security~\cite{sabelfeld2003language}, decomposing stack safety into
the {\em integrity} and {\em confidentiality} of the caller’s local state
during the callee's execution, and the confidentiality of the calle's local state
afterward, as well as the control flow protection of {\em well-bracketed control
  flow (WBCF)}~\cite{SkorstengaardSTKJFP}. We extend WBCF to a setting with exceptions,
in which there may be multiple points at which a callee can ``return.''

A key technical novelty in these definitions,
compared to standard formulations of confidentiality and integrity from the
security literature, is that they
are ``nested'': {\em each} caller is guaranteed protection from its
immediate callees (which, in turn, need protection from their immediate
callees, etc.).
Caller confidentiality is especially interesting. It is based on a traditional
notion of noninterference, but whereas ordinary noninterference
is an end-to-end hyper-property on whole program runs, caller
confidentiality is a nested form of noninterference that applies to
part of a run, requiring that the callee’s behavior is
invariant under hypothetical scrambling of the caller’s stack
but allowing the caller to access its own data upon return.

Our threat model is very strong, allowing the attacker
to execute arbitrary code, including attempting to smash
the stack to disrupt the program’s control flow. Our stack safety
properties demand that, even in the presence of such attackers, confidentiality
and integrity still apply.
%

For confidentiality and integrity, our properties are {\em observational},
meaning that they are only concerned with the observable behavior of the system.
A function activation might read from and write to other activations' data
without violating an observational property, as long as such reads and writes do not
affect the observable behavior of the
system. This permissiveness enables our properties to handle lazy enforcement mechanisms.

To demonstrate the utility of our formal characterization, we use these
properties to validate and improve an existing enforcement mechanism, the
{\em stack safety micro-policies} of Roessler and DeHon~\cite{DBLP:conf/sp/RoesslerD18}, re-implemented
in the Coq proof assistant on top of a RISC-V specification.  We
use QuickChick~\cite{Denes:VSL2014,Pierce:SF4}, a property-based testing
tool for Coq, to generate random programs and check
that Roessler and DeHon's micro-policies correctly abort the ones that
attempt to violate one of our properties. Furthermore, we
%
check that the testing framework is able to generate counterexamples
that violate our properties but are \emph{not} halted by incorrect
enforcement variants---both variants that we accidentally created
during our re-implementation of the micro-policy and ones that we
intentionally crafted to be broken in order to increase our confidence
in testing and the enforcement mechanism itself.

@@

\begin{itemize}
\item We formalize permissive stack confidentiality and integrity and integrity properties.
  Both are parameterized over a notion of external
  observation, and are violated only if accessing secrets or overwriting
  data causes a visible change in the system behavior %(\cref{sec:lazy}).
\item We apply these definitions to a complex system with features such as argument passing
  on the stack, callee-saves registers, exceptions, and a simple coroutine system% (\cref{sec:ext}).
\item We use property-based random testing to validate the relationship between
  our properties and micro-policies %(\cref{sec:testing})
\end{itemize}
%\Cref{sec:relwork,sec:future} discuss related and future work.

Our artifact contains Coq formalizations of our properties for illustrative purposes
as well as the testing implementation.

@@

  We state our properties in terms of a generic machine, making minimal assumptions
  on the machine architecture and the nature of its built-in safety enforcement mechanisms
  (if any).  Our machine model is meant to be a faithful description of the hardware,
  with a semantics that covers both ordinary instruction behavior and the behavior of
  any tags, capabilities, etc. that the machine might support. 
  This semantics is intentionally \emph{not} ``safe-by-construction;'' it gives
  no guarantee that the enforcement hardware is being used correctly to achieve desired
  security properties.

@@

  The security-relevant operations typically correspond to underling machine instructions,
  but the correspondence may not be obvious just be inspecting the machine code.
  For example, in the tagged RISC-V machine we use in our examples and tests,
  calls and returns are conventionally performed by {\tt jal} (``jump-and-link'')
  and {\tt jalr} (``jump-and-link-register'') instructions, respectively, but these
  instructions might also be used for other things. We therefore assume that there
  is some external mechanism for distinguishing which instructions are security-relevant;
  the most obvious technique is to add optional labels to instructions, and that is 
  what we do in this paper, but the theory does not depend on this choice.

@@

In the next section, we give a detailed example of a security semantics for a simple setting
in which our security-relevant operations are restricted to making calls and returns, and allocating
private memory within the current function activation. In later sections, we extend this model
to the full system that we test, which features:
\begin{itemize}
\item Function calls and returns, with caller- and callee-saved registers
\item Allocation of strictly private and strictly public regions on the stack
\item Arguments spilled onto the stack and/or passed by reference
\item Exceptions
\item Tail-call Elimination
\end{itemize}

Finally, we give a property definition for stack safety in which (some) stack-allocated
objects are governed by a provenance-based capability system, accessible only to those
functions which have received a valid pointer to them.

@@

In general, it is impossible to distinguish buggy source code from an attacker; in
our examples we will identify one function or another as an attacker, but we do not
require any static division between trusted and untrusted code, and we aim to protect
even buggy code.

This is a strong threat model, but hardware and timing attacks are out of scope,
and our properties are termination insensitive as a result of how failstops are
modelled (see ~\ref{??}).

@@

Our concurrency model is fairly
simplistic, assuming a fixed number of threads each with its own dedicated processor.
\sna{There is no concurrency in this document. Attempting to build it in from the beginning
  was rather complicated. The question is where to put that extra complexity? It's not difficult
  from a technical standpoint, subject to the one-thread-per-processor constraint.}
We model memory-safe stack objects, but not a heap. Regions outside of
stacks can be used however the compiler likes, including as a heap, but no protection is
built in and our properties assume that if a pointer to a stack object is stored there,
it is permanently compromised.

@@

A {\em machine state} \(\mach\) is a map from elements to a set \(\mathcal{V}\) of
\emph{values}. Each value \(v\) contains a \emph{payload} word, written \(|v|\).
The details of value structure depend on the specific machine being modelled;  
intuitively, the payload represents the part of the value that is relevant to
the behavior of the basic machine, while the rest of the value may contain
information relevant to a hardware enforcement mechanism (such as a tag).
We write \(\mach[\component]\) to denote the value of \(\mach\) at
\(\component\)  and \(\mach[v]\) as shorthand for \(\mach[|v|]\).

{\em Events} are drawn from a set \(\OBSS\) and ranged over by \(\obs\), with a
special silent event \(\tau\).
The machine step function
\(\mach \xrightarrow{\overline{\psi},\obs} \mach' \in \MACHS \rightarrow
\MACHS \times \mathit{list} ~ \Psi \times \OBSS\)
is labeled by an ordered list of security-relevant operations, defined below,
and an event.

While some enforcement mechanisms may cause a machine to
failstop, we model this as the machine silently diverging by remaining in the
same state. We consider silently diverging traces to be indistinguishable
(from the point of view of an attacker) from traces that might eventually
produce output, an assumption that gives our properties the flavor of
{\it termination insensitivity}.

\subsection{Security-Relevant Operations, Principals, and Security State}

Security-relevant operations consist of calls, returns, and allocations.
Calls identify the address of the call's target,
and the argument registers. Allocations and deallocations are phrased in terms of a pair
of an offset from the stack pointer and a size. Returns don't require any additional information.
%
\begin{align*}
  \psi \in \Psi ::= & \mathbf{call} ~ \addr_{target} ~ \overline{\reg_{args}} & \\
  | & \mathbf{return} & \\
  | & \mathbf{alloc} ~ \mathit{off ~ sz} & \mathit{off, sz} \in \mathbb{N} \\
  | & \mathbf{dealloc} ~ \mathit{off ~ sz} & \mathit{off, sz} \in \mathbb{N} \\
\end{align*}
%
Security principals are function activations: at any given time, pending activations
have security requirements that are recorded in a call stack, and the current
activation tracks context information that will inform its requirements when
it makes a call. Each activation has a {\it view}
of the system that maps each state element to a {\it security class}:
it may be private to an inactive principal (\(\sealed\)),
in an allocated object that is accessible to the current principal (\(\object\)),
available to be allocated (\(\unsealed\)), or
always accessible (\(\public\)). The {\it initial view} \(V_0\) maps all stack locations to \(\unsealed\),
all other locations to \(\public\), and registers based on which set they
belong to: % (Figure \ref{fig:RISCVregs}):
\(\sealed\) for callee-saved, \(\unsealed\) for caller-saved, and \(\public\) otherwise.
%Elements of \(\mathcal{P}\),
%such as the program counter, are always \(\public\). Even if they might deserve their
%own protection, that protection is orthogonal to stack safety.

@@

In this section we consider a simple setting with calls, returns, and private allocations,
on a RISC-V machine with a PIPE tagging system. Here 
\(\mathcal{V} ::= \mathcal{W} \times \mathcal{T}\),
for some type \(\mathcal{T}\) of tags to be elaborated later, with \(|(w,t)| = w\).
Figure \ref{fig:RISCVregs} gives an example division of the RISC-V registers into
security classes. 
Figure \ref{fig:main} gives the C code and corresponding compiled RISC-V code
for a function {\tt main}. (Code is compiled using the standard RISC-V
calling conventions~\cite{??}; the stack grows towards lower addresses. We assume that
{\tt main} begins at address 0 and its callee {\tt f} at address 100.)

@@

Suppose that {\tt f} is not a (valid) C function at all, but an attacker seeking
to leak {\tt secret}! It might do so in a number of ways, shown as snippets of
assembly code in Figure \ref{fig:f}.\footnote{We arrange that all the snippets
have the same length just to simplify some later explanations.}
%
In Figure \ref{subfig:direct}, {\tt f} takes an offset from the stack
pointer, accesses {\tt secret}, and directly outputs it. But more
subtly, even if somehow prevented from outputting {\tt secret} directly, {\tt f}
can instead return that value so that {\tt main} stores it to {\tt out},
as in Figure \ref{subfig:indirect}.
%
Beyond simply reading {\tt secret}, the attacker might overwrite {\tt sensitive}
with 42, guaranteeing that {\tt main} publishes its own secret unintentionally
(Figure \ref{subfig:integrity})!
Attacks of this kind do not violate {\tt main}'s confidentiality, but its
{\it integrity}.
In Figure \ref{subfig:WBCF}, the attacker arranges to return to the
wrong instruction, thereby bypassing the check and publishing {\tt secret} regardless,
violating the program's {\it well-bracketed control flow} (WBCF.) Figure \ref{subfig:WBCF2}
shows another attack that violates WBCF, this time by returning to the correct
program counter but with the stack pointer offset.

@@

To see why WBCF also cares about the stack pointer, consider the attack in
Figure \ref{subfig:WBCF2}. Here the attacker returns with \(\SP' = 998\) instead of the
correct \(\SP = 980\). In this scenario, given the layout of {\tt main}'s frame 
\begin{tabular}{| l | l | l | l | l |}
  \multicolumn{1}{r}{\(\SP \downarrow\)} &
  \multicolumn{2}{r}{\(\SP' \downarrow\)} \\
  \hline
  res & sens & sec & ra & ra \\
  \hline
\end{tabular}

@@

\noindent
{\tt main}'s attempt to read {\tt sensitive} will instead 
read part of the return address, and its attempt to output
{\tt res} will instead output {\tt secret}! Even absent other
kinds of data protection, the stack pointer {\it must} be restored
for the program to behave predictably.

@@

To perform this enforcement using tags, we can take
\(\mathcal{T} ::= \mathit{gen} | \mathit{bseq} ~ n ~ m | \mathit{prot}\), named for general
execution, blessed sequence, and protected, respectively. Blessed sequences of length \(m\) are
tagged \(\mathit{bseq} ~ 1 ~ m\) through \(\mathit{bseq} ~ m ~ l\), and the micro-policy enforces
that they execute in order. The stack pointer and {\tt ra}
are tagged \(\mathit{prot}\). Only instructions tagged \(\mathit{bseq}\) may write to memory
or registers tagged \(\mathit{prot}\). Store instructions tagged \(\mathit{bseq}\) tag their
targets \(\mathit{prot}\).

@@

While it would be simple to define integrity as ``all sealed elements retain their
values after the call,'' this would be stricter than necessary for many purposes.
For example, it arguably doesn't matter if {\tt f} overwrites {\tt main}'s data provided
that {\tt main} never reads that corrupted data.
To illustrate, we can consider two integrity micro-policies: ``Depth Isolation'' and
``Lazy Tagging and Clearing.''

Depth Isolation tracks the depth of the current activation via a tag on the program counter.
Stack frames are initialized and cleared, by writing
to every address in them, so that during a call at depth \(d\), the current activation's
frame is tagged with \(d\). Loads and stores targeting addresses in frames failstop
if the program counter tag does not match the frame tag. In this setting, at depth \(d\),
all memory tagged with a depth less than \(d\) corresponds to \(\sealed\) memory,
and should be indeed be inaccessible.

@@

In our example setting, the observation trace consists of the sequence of values written to {\tt out}.
The example in Figure \ref{subfig:integrity} modifies the value of {\tt sensitive},
which is \(\sealed\). Figure \ref{fig:variant} shows the state just after the call at step 5,
assuming that {\tt sec} is 5. Similar to WBCF, we will define
\(\intProp\) as a predicate on states that holds if
all relevant sealed addresses in \(\mach\) are the same as after step 5.
We once again require that \(\intProp\) hold on the state following the matching return, 
which is reached by step 9. Here {\tt sensitive} has obviously changed, but is it relevant?
Consider a variant state in which {\tt sensitive} has any other value, arbitrarily
choosing 43 (and keeping its tag, \(t_2'\), unchanged.) As execution continues after the return
from the original state, it passes the {\tt if}-test on {\tt sensitive}, whereas the execution
from the variant does not, resulting in differing outputs. Therefore {\tt sensitive} \emph{is} relevant,
so \(\intProp\) does not hold, and integrity has indeed been violated.

@@

Others combine this protection with some degree of memory protection,
chiefly focusing on integrity. Skorstengaard et al. \cite{SkorstengaardSTK} characterize stack
safety as the combination of WBCF and {\it local state encapsulation} (LSE), and prove that their
capability-based enforcement satisfies both. LSE corresponds to integrity, plus the
confidentiality of the caller's frame, but not of the unallocated stack or data left
behind in it by previous callees.
They offer a fully abstract overlay semantics, which formalizes WBCF and LSE
together as a safe-by-construction abstract machine. That machine does not allow a callee to
read or write its caller's frame, but might allow it to read data that is left behind below
the caller's frame, which is weaker than our full caller confidentiality and rules out
callee confidentiality entirely. Georges et al. \cite{Georges22:TempsDesCerises} correct
both limitations.

@@

\subsection{Facts Abouts Calls and Returns}
\label{sec:facts}

>>>>

The term {\em stack safety} is associated with a
variety of compiler, run-time, and hardware mechanisms for protecting stack
memory. But, unlike ``the heap,'' the stack does not correspond to a single
high-level language concept, and mechanisms for protecting it
typically lack precise specifications,
relying instead on informal descriptions and examples of bad
behaviors that they prevent. A more rigorous specification is especially desirable
for unusual enforcement mechanisms\bcp{why especially for this?}
like the ``lazy'' stack safety
micro-policies proposed by Roessler and DeHon~\cite{DBLP:conf/sp/RoesslerD18}.
% The term {\em Stack safety} is associated with a
% variety of compiler, run-time, and hardware mechanisms for protecting stack
% memory. But unlike heap memory, the stack does not correspond to any single
% high-level language concept. The result is that
% these mechanisms typically lack precise specifications,
% relying instead on informal descriptions and examples of the bad
% behaviors that they prevent. Such a specification is especially desirable
% for an unusual enforcement mechanism, like the ``lazy'' stack safety
% micro-policies proposed by Roessler and DeHon~\cite{DBLP:conf/sp/RoesslerD18}.

We propose a formal characterization of ``observable'' stack safety based on
concepts from language-based security. Stack safety is a
combination of an integrity property (``the private
state in each caller's stack frame is held invariant by the callee'')
and\bcp{two?} confidentiality properties\bcp{plural?} (``the caller's
and callee's behavior are insensitive to one another's private
state,''), both compatible with lazy tagging\bcp{undefined term; also,
the fact that we've mentioned tagged enforcement twice already makes
it really feel like we are proposing an abstract theory of exactly one
thing}, plus a control-flow property.

@@

The call stack is a perennial target for low-level attacks, leading to
dire consequences from leakage and corruption of private stack data
to control-flow hijacking. To foil such attacks, a menagerie of
software and hardware protections have been proposed,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges+21},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18}.
  \ifaftersubmission\apt{Mostly from
  nick; there could be more}\bcp{Yes, going back to MIT days---we should
  include several more of these, if only to give readers the impression that
this is a well-studied mechanism (so formalizing its protections is
useful).}
\fi

The protections offered by such mechanisms are commonly described by giving
concrete examples of attacks that they can prevent---corruption of return
addresses, buffer overflows, use of uninitialized variables, etc.---leaving
a more generic characterization to the reader's intuition.  But these
mechanisms can be quite intricate, and the behaviors they aim to prevent are
subtle and varied. It can be hard to guess, based only on informal
intuitions, whether a given mechanism actually blocks all potential
attacks in a given class---or, conversely, whether it is overly conservative and disallows
system behaviors that are actually safe.
To settle such questions with confidence,
we need a precise, generic, and formal specification for stack
safety, as a basis both for comparing the security claims of different
enforcement techniques and for validating that these claims
are met by particular implementations. With the notable exception of the model used
in Skorstengaard et al.'s StkTokens work~\cite{SkorstengaardSTKJFP}, which we
discuss in Section \ref{sec:relwork}, there is a dearth of such properties in
the literature. \bcp{can we give a very quick preview here of the comparison?}

We propose a novel characterization of stack safety using the formal tools of language-based
security~\cite{sabelfeld2003language}, decomposing stack safety into
the {\em integrity} and {\em confidentiality} of the caller’s local state
during the callee's execution, the confidentiality of the callee's local state
afterward, and the control flow protection of {\em well-bracketed control
  flow (WBCF)}~\cite{SkorstengaardSTKJFP}. We also extend WBCF to a setting with exceptions,
in which there may be multiple points at which a callee can ``return.''

A key technical novelty in these definitions,
compared to standard formulations of confidentiality and integrity from the
security literature, is that they
are ``nested'': {\em each} caller is guaranteed protection from its
immediate callees (which, in turn, are protected from their immediate
callees, etc.).
Caller confidentiality is especially interesting: It is based on a traditional
notion of noninterference, but, whereas ordinary noninterference
is an end-to-end hyper-property on whole program runs, caller
confidentiality is a nested form of noninterference that applies to
part of a run, requiring that the callee’s behavior be
invariant under hypothetical scrambling of the caller’s stack
but allowing the caller to access its own data upon return.

Our threat model is very strong: it allows the attacker
to execute arbitrary code, including attempting to smash
the stack to disrupt the program’s control flow. Our stack safety
properties promise that, even in the presence of such attackers, confidentiality
and integrity still apply.
%

For confidentiality and integrity, our properties\bcp{by this point,
  the reader may be getting confused about why we are talking about
  multiple properties, rather than just one (with multiple clauses)} are {\em observational},
meaning that they are only concerned with the observable behavior of the system.
A function activation might read from and write to other activations' data
without violating an observational property, as long as such reads and writes do not
affect the observable behavior of the
system. This permissiveness enables our properties to handle lazy
enforcement mechanisms.  \bcp{do we have any in mind aside from Nick's
tagging-based ones?}

To demonstrate the utility of our formal characterization, we use these
properties to validate and improve an existing enforcement mechanism, the
{\em stack safety micro-policies} of Roessler and DeHon~\cite{DBLP:conf/sp/RoesslerD18}, re-implemented
in the Coq proof assistant on top of a RISC-V specification.  We
use QuickChick~\cite{Denes:VSL2014,Pierce:SF4}, a property-based testing
tool for Coq, to generate random programs and check
that Roessler and DeHon's micro-policies correctly detect the ones that
attempt to violate one of our properties. Furthermore, we
%
check that the testing framework is able to generate counterexamples
that violate our properties but are \emph{not} halted by incorrect
variants of the enforcement mechanisms, including ones that we accidentally created
while implementating the micro-policy and ones that we
intentionally crafted to increase our confidence in the effectiveness
of our testing.

@@

\begin{itemize}
\item We formalize permissive stack confidentiality and integrity properties.
  Both are parameterized over a notion of external
  observation, and are violated only if accessing secrets or overwriting
  data causes a visible change in system behavior %(\cref{sec:lazy}).
\item We apply\bcp{instantiate?} these definitions to a complex\bcp{is
  its complexity a selling point?} system
  with features such as argument passing
  on the stack, callee-saves registers, exceptions, and a simple
  coroutine system% (\cref{sec:ext}).
  \bcp{Are these features really independent of the generic model?  If
  so, that's a nice point that should be brought out clearly.}
\item We use property-based random testing to validate the relationship between
  our properties and micro-policies. %(\cref{sec:testing})
  \bcp{Micro-policies have not been explained, and we haven't
    motivated why this would be an important or useful thing to do.}
\end{itemize}
%\Cref{sec:relwork,sec:future} discuss related and future work.
\bcp{Why did the section references get taken away from all that?  And
why delete the last sentence?}

Our artifact contains Coq formalizations of our properties for illustrative purposes
as well as the testing implementation.
\bcp{Make clear that there are no proofs!}

@@

  We state our properties in terms of a generic machine, making minimal assumptions
  on the machine architecture and the nature of its built-in safety enforcement mechanisms
  (if any).  Our machine model is meant to be a faithful description of the\bcp{??} hardware,
  with a semantics that covers both ordinary instruction behavior and the behavior of
  any tags, capabilities, etc. that the machine might support.
  This semantics is intentionally \emph{not} ``safe-by-construction;'' it gives
  no guarantee that the enforcement hardware is being used correctly to achieve desired
  security properties.

@@

  The security-relevant operations typically correspond to underling machine instructions,
  but the correspondence may not be obvious just by inspecting the machine code.
  For example, in the tagged RISC-V machine we use in our examples and tests,
  calls and returns are conventionally performed by {\tt jal} (``jump-and-link'')
  and {\tt jalr} (``jump-and-link-register'') instructions, respectively, but these
  instructions might also be used for other things. We therefore assume that there
  is some external mechanism for distinguishing which instructions are security-relevant;
  the most obvious technique is to add optional labels to instructions, and that is
  what we do in this paper, but the theory does not depend on this choice.

@@

In the next section, we give a detailed example of a security semantics for a simple setting
in which our security-relevant operations are restricted to calls,
returns, and allocation of
private memory within the current function activation. In later sections, we extend this model
to the full system that we test, which features
% \begin{itemize}
% \item
function calls and returns, with caller- and callee-saved registers,
% \item
allocation of strictly private and strictly public regions on the stack,
% \item
arguments spilled onto the stack and/or passed by reference,
% \item
exceptions,
and
% \item
tail-call elimination.
% \end{itemize}

Finally, we give a property definition for stack safety in which (some) stack-allocated
objects are governed by a provenance-based capability system, accessible only to those
functions which have received a valid pointer to them.  \bcp{???}

@@

In general, it is impossible to distinguish buggy machine code from an attacker; in
our examples we will identify one function or another as an attacker, but we do not
require any static division between trusted and untrusted code, and we aim to protect
even buggy code.

This is a strong threat model, but it does omit some important aspects of
security enforcement: In particular, hardware and timing attacks are out of scope,
and our properties are termination insensitive as a result of how failstops are
modelled (see~\ref{??}).

@@

Our concurrency model is fairly
simplistic, assuming a fixed number of threads each with its own dedicated processor.
\sna{There is no concurrency in this document. Attempting to build it in from the beginning
  was rather complicated. The question is where to put that extra complexity? It's not difficult
  from a technical standpoint, subject to the one-thread-per-processor constraint.}\bcp{Appendix?}
We model memory-safe stack objects, but not a heap. Regions outside of
stacks can be used however the compiler likes, including as a heap, but no protection is
built in and our properties assume that if a pointer to a stack object is stored there,
it is permanently compromised.

@@

A {\em machine state} \(\mach\) is a map from state elements to a set \(\mathcal{V}\) of
\emph{values}. Each value \(v\) contains a \emph{payload} word, written \(|v|\).
The details of value structure depend on the specific machine being modelled;
intuitively, the payload represents the part of the value that is relevant to
the behavior of the basic machine, while the rest of the value may contain
information relevant to a hardware enforcement mechanism (such as a tag).
We write \(\mach[\component]\) to denote the value of \(\mach\) at
\(\component\)  and \(\mach[v]\) as shorthand for \(\mach[|v|]\).

{\em Events} are drawn from a set \(\OBSS\)\bcp{lowercase?} and ranged over by \(\obs\), with a
special silent event \(\tau\).
The machine step function
\(\mach \xrightarrow{\overline{\psi},\obs} \mach' \in \MACHS \rightarrow
\MACHS \times \mathit{list} ~ \Psi \times \OBSS\)
is labeled by an ordered list of security-relevant operations, defined below,
and an event.

While some enforcement mechanisms may cause a machine to
failstop, we model this as the machine silently diverging by remaining in the
same state. We consider silently diverging traces to be indistinguishable
(from the point of view of an attacker) from traces that might eventually
produce output\bcp{why?}, an assumption that gives our properties the flavor of
{\it termination insensitivity}.

\subsection{Security-Relevant Operations, Principals, and State}

Security-relevant operations consist of calls, returns, and allocations.
Calls identify the address of the call's target,
plus the argument registers. Allocations and deallocations are phrased in terms of a pair
of an offset from the stack pointer and a size. Returns don't require any additional information.
%
\begin{align*}
  \psi \in \Psi \ \ ::= \ \ & \mathbf{call} ~ \addr_{target} ~ \overline{\reg_{args}} & \\
  | \ & \mathbf{return} & \\
  | \ & \mathbf{alloc} ~ \mathit{off ~ sz} & \mathit{off, sz} \in \mathbb{N} \\
  | \ & \mathbf{dealloc} ~ \mathit{off ~ sz} & \mathit{off, sz} \in \mathbb{N}
\end{align*}
%
The security principals in our account of stack safety are function
activations: at any given time, pending activations
have security requirements that are recorded in a call stack, and the current
activation tracks context information that will inform its requirements when
it makes a call. Each activation has a {\it view}
of the system that maps each state element to a {\it security class}:
it may be private to an inactive principal (\(\sealed\)),
in an allocated object that is accessible to the current principal (\(\object\)),
available to be allocated (\(\unsealed\)), or
always accessible (\(\public\)). The {\it initial view} \(V_0\) maps all stack locations to \(\unsealed\),
all other locations to \(\public\), and registers based on which set they
belong to: % (Figure \ref{fig:RISCVregs}):
\(\sealed\) for callee-saved, \(\unsealed\) for caller-saved, and \(\public\) otherwise.
%Elements of \(\mathcal{P}\),
%such as the program counter, are always \(\public\). Even if they might deserve their
%own protection, that protection is orthogonal to stack safety.

@@

In this section we consider a simple setting with calls, returns, and private allocations,
on a RISC-V machine with a PIPE tagging system\bcp{undefined}. Here
\(\mathcal{V} ::= \mathcal{W} \times \mathcal{T}\),
for some type \(\mathcal{T}\) of tags to be elaborated later, with \(|(w,t)| = w\).
Figure \ref{fig:RISCVregs} gives an example division of the RISC-V registers into
security classes.
Figure \ref{fig:main} gives the C code and corresponding compiled RISC-V code
for a function {\tt main}. Code is compiled using the standard RISC-V
calling conventions~\cite{??}; the stack grows towards lower addresses. We assume that
{\tt main} begins at address 0 and its callee {\tt f} at address 100.

@@

Suppose that {\tt f} is actually an attacker seeking
to leak {\tt secret}. It might do so in a number of ways, shown as snippets of
assembly code in Figure \ref{fig:f}.\footnote{We arrange that all the snippets
have the same length just to simplify some later explanations.}
%
In Figure \ref{subfig:direct}, {\tt f} takes an offset from the stack
pointer, accesses {\tt secret}, and directly outputs it. But more
subtly, even if somehow prevented from outputting {\tt secret} directly, {\tt f}
can instead return that value so that {\tt main} stores it to {\tt out},
as in Figure \ref{subfig:indirect}.
%
Beyond simply reading {\tt secret}, the attacker might overwrite {\tt sensitive}
with 42, guaranteeing that {\tt main} publishes its own secret unintentionally
(Figure \ref{subfig:integrity}).
Attacks of this kind do not violate {\tt main}'s confidentiality, but
rather its {\it integrity}.
In Figure \ref{subfig:WBCF}, the attacker arranges to return to the
wrong instruction, thereby bypassing the check and publishing {\tt secret} regardless,
violating the program's {\it well-bracketed control flow} (WBCF.) In
Figure \ref{subfig:WBCF2}, a different attack violates WBCF, this time
by returning to the correct program counter but with the wrong stack pointer.

@@

To see why WBCF also cares about the stack pointer, consider the attack in
Figure \ref{subfig:WBCF2}. Here the attacker returns with \(\SP' = 998\) instead of the
correct \(\SP = 980\). In this scenario, given the layout of {\tt main}'s frame,
\begin{center}
\begin{tabular}{| l | l | l | l | l |}
  \multicolumn{1}{r}{\(\SP \downarrow\)} &
  \multicolumn{2}{r}{\(\SP' \downarrow\)} \\
  \hline
  res & sens & sec & ra & ra \\
  \hline
\end{tabular}
\end{center}

@@

\noindent
{\tt main}'s attempt to read {\tt sensitive} will instead
read part of the return address, and its attempt to output
{\tt res} will instead output {\tt secret}!
% Even absent other
% kinds of data protection, the stack pointer {\it must} be restored
% for the program to behave predictably.

@@

To perform this enforcement using tags, we can take
\(\mathcal{T} ::= \mathit{gen} \mathrel{|} \mathit{bseq} ~ n ~ m \mathrel{|} \mathit{prot}\), named for general
execution, blessed sequence, and protected, respectively. Blessed sequences of length \(m\) are
tagged \(\mathit{bseq} ~ 1 ~ m\) through \(\mathit{bseq} ~ m ~ l\), and the micro-policy enforces
that they execute in order. The stack pointer and {\tt ra}
are tagged \(\mathit{prot}\). Only instructions tagged \(\mathit{bseq}\) may write to memory
or registers tagged \(\mathit{prot}\). Store instructions tagged \(\mathit{bseq}\) tag their
targets \(\mathit{prot}\).

@@

While it would be simple to define integrity as ``all sealed elements retain their
values after the call,'' this would be stricter than necessary for many purposes.
For example, it arguably doesn't matter if {\tt f} overwrites {\tt main}'s data provided
that {\tt main} never reads that corrupted data.
To illustrate why this matters, let's consider two integrity
micro-policies: ``Depth Isolation'' and
``Lazy Tagging and Clearing.''

Depth Isolation tracks the depth of the current activation via a tag on the program counter.
Stack frames are initialized and cleared by writing
to every address in them, so that during a call at depth \(d\), the current activation's
frame is tagged with \(d\). Loads and stores targeting addresses in frames failstop
if the program counter tag does not match the frame tag. In this setting, at depth \(d\),
all memory tagged with a depth less than \(d\) corresponds to \(\sealed\) memory,
and should be indeed be inaccessible.

@@

In our example setting, the observation trace consists of the sequence
of values written to {\tt out}.\bcp{Is this the first place we've
  discussed what the observations actually are?  Is this the only
  thing they could be, or does the model support other options?}
The example in Figure \ref{subfig:integrity} modifies the value of {\tt sensitive},
which is \(\sealed\). Figure \ref{fig:variant} shows the state just after the call at step 5,
assuming that {\tt sec} is 5. Similar to WBCF, we will define
\(\intProp\) as a predicate on states that holds if
all relevant sealed addresses in \(\mach\) are the same as after step 5.
We once again require that \(\intProp\) hold on the state following the matching return,
which is reached by step 9. Here {\tt sensitive} has obviously changed, but is it relevant?
Consider a variant state in which {\tt sensitive} has any other value, arbitrarily
choosing 43 (and keeping its tag, \(t_2'\), unchanged.) As execution continues after the return
from the original state, it passes the {\tt if}-test on {\tt sensitive}, whereas the execution
from the variant does not, resulting in differing outputs. Therefore {\tt sensitive} \emph{is} relevant,
so \(\intProp\) does not hold, and integrity has indeed been violated.

@@

Others combine this protection with some degree of memory protection,
chiefly focusing on integrity. Skorstengaard et al. \cite{SkorstengaardSTK} characterize stack
safety as the combination of WBCF and {\it local state encapsulation} (LSE), and prove that their
capability-based enforcement satisfies both. LSE corresponds to integrity, plus the
confidentiality of the caller's frame, but not of the unallocated stack or data left
behind in it by previous callees.
They offer a fully abstract overlay semantics, which formalizes WBCF and LSE
together as a safe-by-construction abstract machine. That machine does not allow a callee to
read or write its caller's frame, but might allow it to read data that is left behind below
the caller's frame, which is weaker than our full caller confidentiality and rules out
callee confidentiality entirely. Georges et al. \cite{Georges22:TempsDesCerises} correct
both limitations.\bcp{Then why are we not comparing to them instead??}

@@

\subsection{Facts About Calls and Returns}
\label{sec:facts}
